Assessing paravascular transport in the parenchyma by PDE constrained optimization.

### 2018-10-23 tex/paper.tex
update

\newcommand{\kam}[1]{\textcolor{blue}{#1}}

@@

\begin{abstract}
bla bla 
\end{abstract}
\section{Introduction}
Recent breakthroughs in medicine have linked neurodegenerative diseases like Alzheimer's to insufficient sleep through a novel fluid mechanism called the glymphatic system 
that is responsible for waste clearance. 
In 2011,~\citet{iliff2012paravascular} discovered a novel system within our brain's metabolism system, which changed our understanding of the brain. 
The system involves a new pathway, called the paravascular pathway as it runs in parallel with vasculature system in a surrounding compartment.
The important novelty in the 2011 findings was that waste from the brain is cleared through the paravascular path, and that a malfunction of this system 
is linked to the development of dementia such as Alzheimer's and Parkinson's diseases. In 2013,~\citet{xie2013sleep} found
that this system is particularly active during sleep.
These mentioned breakthroughs were all performed in mice. In 2018,  
it was for the first time demonstrated the action of
this system in humans, and that the action of the system differs
between healthy controls and patients with dementia~\citet{ringstad2018brain}. 
However, the biomechanical mechanisms behind this is not well understood from a fluid mechanics perspective,
and so far the modeling attempts have mostly failed~\citet{asgari2016glymphatic, holter2017interstitial, smith2017glymphatic}. 
In particular the system seems to facilitate waste transport faster than  
the diffusion, which since the pioneering works of \citet{sykova2008diffusion} 
has been the prevailing paradigm. Here, we aim to investigate whether PDE constrained optimization can be used
to assess the effective diffusion coefficient on long time-scales (hours or days).

@@

In \cite{ringstad2018brain} brain-wide distribution of MRI-contrast was demonstrated during 24 hours after lumbar contrast injection. Brain-wide distribution by diffusion alone was deemed unlikely by the authors (some of which are co-authors). The argument was based on analytical considerations where it was calculated that 50\% contrast enrichment would occur after 
55 hours using the error function which is valid for planar diffusion. However, 
the surface of the brain is folded and is around five times larger than 
a corresponding surface of a ball with the same volume. Hence, 
a more rigorous modeling attempt is warranted. A complicating factor
is however that the contrast in the surrounding CSF is heterogenous
and changes significantly during the 24 hours of the investigations. 
Furthermore, images were obtained only at a few time-points during the investigations in \cite{ringstad2018brain}. As such the sparseness i times prevent a direct computation of the diffusion coefficients.

The pioneering works of Sykov{\'a} and Nicholson~\cite{sykova2008diffusion} demonstrated
that diffusion was a governing transport mechanism in the brain and estimated to be on the order
of $1.0e-4 \mathrm{mm\sp{2}/s}$ for large molecules. This is confirmed with DTI where values of young and healthy are typically around $0.7e-3 \mathrm{mm\sp{2}/s}$ \cite{Helenius194}
while values in dementia typically are XXX \kam{Geir vet du dette?}. Of course diffusion coefficient 
measured by DTI at a time-scale of 7-30 minutes, may not be representative for the process on a
longer time-scale, at least if the paravascular pathway plays an important role. The fact that the vasculature  occupy around 3\% of the brain volume 
and the paravascular network is substatantially smaller potentially render it invisible at the short time-scales of a DTI investigation 
while it may still be effective on longer time scales.

Our purpose in this paper is to attempt a more rigorous methodology 
for assessing the apparent diffusion process where
the complex geometry of the brain is taken into account and time-scales of hours or days are taken into account. 
The approach we will take is to solve a PDE constrained optimization problem 
via adjoint methods where we have sparse observations on selected time-points, around 10 acquisitions during 24 hours. But through
the complete domain. Hence, the challenges faced from a mathematical point of view 
are that 1) the images are subject to noise, 2) resolution in space is limited to slightly less than (1 mm)$^3$ and 3)
the sparse observations through the time domain.  As such we need to assess the sensitivity of
the approach with respect important parameters such as to regularization parameters, noise levels and 
time resolution to determine whether this approach is a viable method
to obtain parameters involved time-scales. We remark that the purpose
of this paper is a systematic study of the mathematical challenges and that assessing  
whether clearance is governed by a process that is faster than diffusion is not the topic of the current paper.

@@

Section 2.1-2.3 describes the details of the imaging relevant for this study. More details can however be
found in e.g. \cite{ringstad2018brain}. Section 2.4-2.5 describe the solution of PDE constrained optimization problems and
its implementation in FEniCS. The exposition is, however, brief and we refer to XXX for more details.

\subsection{MRI Data}
\begin{figure}
\includegraphics[width=0.95\textwidth]{PatID-68-new-100.png} 
\caption{Shows the percentage intensity increase from baseline at different observation times. The colorbar was restricted to the range $(0,100)$ }
\label{fig1} 
\end{figure}

\begin{figure}
\includegraphics[width=0.95\textwidth]{Zoom-PatID-68.png} 
\caption{Shows the percentage intensity increase from baseline at different observation times in the slice (marked red in the left panel) used in the subsequent analysis.The color bar was restricted to the range $(0,100)$}
\label{fig2} 
\end{figure}
Figure~\ref{fig1} shows the distribution of tracers during 24 hours, see also~\cite{ringstad2018brain} for further information on the imaging procedure.   
Our data (not all shown) consist of a total of 10 MRI observations, including a baseline MRI taken before tracer was injected. The observation points are distributed with 5 observations within 1-2 hours after injection, a single observation in the timeframes 2-4 hours, 6-9 hours, 24 hours and 48 hours. 
Figure~\ref{fig2} shows the region selected for our computations. 
The software Freesurfer \cite{Dale1999179, FischlLiuDale, spf2007, reuter:robreg10} was used to segment and align each of the observations, which made it possible to estimate voxelwise intensity increase.

@@

concentration. We remark that we use a notation common in medical literature and here two letter symbols are common. Hence, below we will use two letter symbols such as $TE$ and $TR$ to keep the notation consistent the presentation in \cite{GOWLAND, MPRAGE}.   

@@

The superscripts indicating with contrast $T_{1}\sp{c}$ and without contrast $T_{1}\sp{0}$ and $r_1$ as the relaxivity constant for the MRI tracer in a medium. 

@@

Here $T_b$ is known as the echo spacing, $T_a$ is the inversion time, $T_w$ the time delay, $TR$ as the repetition time, $m$ is the number of echo spacings and $T1$ is the longitudinal(spin-lattice) relaxation time for a given medium. The $M_0$ is a calibration constant of the magnetization. The center echo denoted as $n=m/2$ will be the signal that we will consider when estimating MRI tracer. Given Eq.\ref{EQ::SI} we have that relative intensity increase can be written as 

@@

\frac{S\sp{c}}{S\sp{0}} = \frac{ M_{n}\sp{c} \sin (\theta)}{ M_{n}\sp{0} \sin (\theta) } ,

@@

which can be seen in Figure () . 

@@

and then estimate the concentration using \eqref{EQ::contrast}. The $T_{1}\sp{0}$ values were obtained by T1 mapping the brain using a MRI sequence known as MOLLI5(3)3 \cite{TAYLOR201667}. This takes into account patients specific characteristic, such as tissue damage. Tissue damage can be observed in the MRI due to a lower intensity in the white matter compared to healthy white matter tissue, thus damaged tissue have different $T_1$ relaxation time.

\begin{figure}
\centering
\includegraphics[width=0.70\textwidth]{T1function.png} 
\caption{Show the  function defined in \eqref{scaledmagnetization} where the white region indicates $T_1$ values for white matter,the grey region indicates $T_1$ values for grey matter,the blue region indicates $T_1$ values for CSF.  }
\label{figuredti} 
\end{figure}

\subsection{Diffusion tensor imaging}
In addition to the T1 and T2 weighted sequences described above we have also obtained diffusion tensor imaging (DTI) to assess the apparent diffusion 
coefficients on short time-scales. Images are shown in Figure~\ref{figuredti} where the larges diffusion coefficient (shown in 
red in the middle figure) is shown to be around $1.0e-3 mm^2/s$. We remark that we have not included possible anisotropy, shown in 
the right-most image in Figure~\ref{figuredti} and that these images shows the apparent diffusion coefficient for free water molecules (18 Da).      
The diffusivity of the Gadovist (600 Da \cite{MGadobutrol}) was estimated to be similar to the diffusion coefficient of Gd-DPTA (550 Da \cite{MGgDPTA}). This due to the fact that both molecules have similar mass, and based on Stoke-Einstein equation should have similar diffusion coefficients. The free diffusion coefficient for Gd-DPTA was estimated in \citet{GdDPTA-DIFFUSION} to be $3.8e-4\mathrm{mm\sp{2}/s}$.
The fractional anisotropy is defined as 
\begin{equation}
FA\sp{2} =  \frac{3}{2} \frac{ (\lambda_1 - MD )\sp{2} +(\lambda_2 - MD )\sp{2} +(\lambda_3 - MD )\sp{2}}{\lambda\sp{2}_1 + \lambda\sp{2}_2  +\lambda\sp{2}_3 },
\end{equation}
with the mean diffusivity $MD$ defined as 
\begin{equation}
MD = \frac{\lambda_1 +\lambda_2 +\lambda_3 }{3}.
\end{equation}
In these equations $\lambda_i$ denotes the eigenvalues of the diffusion tensor.
\begin{figure}
\centering
\includegraphics[width=0.80\textwidth]{DTI-zoom.png} 
\caption{The left panel show the anatomical map. The middle panel show the apparent diffusion coefficients (ADC) obtained from DTI. The right panel shows the computed fractional anisotropy (FA) from the DTI.}
\label{figuredti} 
\end{figure}

\subsection{Mathematical Model}
In \cite{sykova2008diffusion}, it was shown that the macroscopic diffusion in the brain can be considered a hindered diffusion with an apparent diffusion coefficient. In order to estimate the apparent diffusion coefficient involved in the tracer transportation shown in Figure~\ref{fig1} we assume that the process can be modeled by a diffusion equation. Then we solve a PDE constrained optimization problem where
the tracer distribution, boundary conditions, and apparent
diffusion coefficient are solved for by optimizing with
respect to the observed tracer distribution. As such, enhanced transportation because of effects such as
dissipation would result in an apparent diffusion coefficient larger than that predicted by DTI.  
The objective function was defined as 
\begin{equation}
\min_{D,g} F = \quad \sum\limits_i\sp{n} \int\limits_{\Omega} |u(t_i) - u_{obs}(t_i)|\sp{2} \mathrm{d}\Omega + \frac{\alpha}{2} \int\limits_{0}\sp{T} || g ||\sp{2}_{L\sp{2}(\partial\Omega)} \mathrm{d}t + \frac{\beta}{2} \int\limits_{0}\sp{T} || \dot{g} ||\sp{2}_{L\sp{2}(\partial\Omega)}\mathrm{d}t 
\label{EQ::objf}
\end{equation}
subject to   
\begin{equation}
\begin{aligned}
\frac{\partial u}{\partial t} = \nabla \cdot  D_i \nabla u \qquad \text{in} \qquad \Omega \times \left\lbrace 0 , T \right)  \\
u=g(t) \qquad \text{on} \qquad \partial\Omega  \times \left\lbrace 0 , T \right) 
\end{aligned}
\label{Eq::PDE}
\end{equation}
Here, $u$ is the contrast distribution, $D_i$ is the apparent diffusion 
coefficient, and $g$ is the boundary condition. Furthermore,  
the domain $\Omega$ contains three sub domains, each with a different diffusion coefficient. We denote the Cerebral Spinal fluid (CSF) domain as $\Omega_1$, the grey matter as $\Omega_2$ and the white matter as $\Omega_3$. The apparent
diffusion constant is assumed to be constant within the CSF, grey and 
white matter but each region may have different values.  
The $\alpha$ and $\beta$ parameters are non-negative regularization parameters 
and $u_{obs}$ are observations at certain time-points $t_i$.

The mesh construction was patient-specific and used the MRI of a patient diagnosed with idiopathic normal pressure hydrocephalus (iNPH). The software Freesurfer was used in segmentation and creating the polyhedral surfaces of the white and grey matter. Then the use of T2 weighted MRI \cite{eidevalnes} was used to segment the CSF compartment surrounding the cerebral and the lateral ventricles. CGAL \cite{cgal:rty-m3-18b} was used to combine the polyhedral surfaces and to construct the mesh. The computational requirement for the resulting mesh was significant, therefore two submesh was also constructed, see Fig.\ref{Fig::Mesh}.
\begin{figure}
\centering
\includegraphics[scale=0.2]{mesh.png} 
\caption{A) Shows the mesh created from the baseline MR image with 3 domains. B) Shows the mesh created from the baseline MR image with 2 domains.  }
\label{Fig::Mesh}
\end{figure}
The 3 domain mesh in Fig.\ref{Fig::Mesh} A, consists of 244318 tetrahedral cells and 22057 vertices, while the 2 domain mesh Fig.\ref{Fig::Mesh} B consists of 335589 tetrahedral cells and 73002 vertices. The dimension of the mesh coordinates are $\mathrm{mm}$.

\subsection{Manufactured Solution}
In order to asses the robustness and accuracy of our strategy and test the dependency of the
regularization parameters, we perform a test case with a
known solution.  
The manufactured observations was obtained by forward computation of Eq.\ref{Eq::PDE} with the Dirichlet boundary condition defined as
\begin{equation}
g(t)_{\partial \Omega_1} = 0.3 +0.167t - 0.007t\sp{2} \qquad  0 \leq t \leq 24.
\label{EQ::DIRI}
\end{equation}
The timestep was $dt = 0.02$, and the diffusion coefficients were selected to be 
\begin{equation}
D_{\Omega_1} = 1000 \quad , \quad D_{\Omega_2} = 4.0 \quad , \quad D_{\Omega_3} = 8.0 
\end{equation}  
The magnitude order for $D_{\Omega_1} $ and $D_{\Omega_2}$ were chosen to resemble diffusion coefficient for water in grey and white matter, but the relation between the coefficients were not preserved. The diffusion coefficient in the CSF was chosen given [cite Bryne and Kent], and the initial condition was set to 0 everywhere. %and the initial condition was set to 1.0 in $\Omega_1$ and 0 elsewhere.
The manufactured forward computation gave a total of 120 possible observation times.

\section{Implementation}
The estimation of the tracer concentration was precomputed and used the parameters obtained from the T1 map, MPRAGE MRI protocol \cite{eidevalnes} and the value for $r_1$ found in \cite{pmid16230904}. In the computation, the function Eq.\ref{scaledmagnetization} was computed for $ T_1\in \lbrace 200, 4000 \rbrace$ creating a lookup table. The lookup table was utilized with the baseline intensity increase to estimate $T_1\sp{c}$, and then the concentration was computed using Eq.\ref{EQ::contrast}.

@@

The code used the FEniCS project (v.2017.2) to solve the PDE, using backwards Euler time discretization and first order continuous Galerkin finite elements. The linear problem was solved using GMRES, and the matrix was stored in the solver. 
The module dolfin adjoint ~\cite{farrell2013automated} \kam{correct ref?}  was used to solve the PDE constrained optimization problem with the L-BFGS-B algorithm in the python module scipy \cite{LBFGSB1, LBFGSB2}. The convergence criteria for the minimization was set so that 
\begin{equation}
||x ||_{L\sp{\infty}} \leq \epsilon
\end{equation} 
with $\epsilon=1.0e-1$ and $x$ denoting the projected gradient.

The observations $u_{obs}$ have a fixed time-points $t_i$, these time-points may not coincide  with the time discretization in the forward problem $t_j$. Therefore a linear interpolation of the current solution $u(t_j)$ and previous solution $u(t_{j-1})$ was used approximate the solution at time $t_i$. The linear interpolation was given as  
\begin{equation}
\label{observation:interpolation}
u(t_i) \approx \frac{\Delta t}{dt} u(t_{j-1}) + \frac{dt - \Delta t }{dt} u(t_{j}), \quad t_i \in \lbrace t_{j-1}, t_j \rbrace
\end{equation}
with $\Delta t = t_j-t_i$ and $dt  = t_{j} - t_{j-1}$.

The implementation used the first observation as initial conditions of Eq.\ref{Eq::PDE}. Then for each time-step, the next observation was used as the initial value for boundary control $g$, which was imposed on the external boundary of the domain $\Omega_1$. In order to increase the convergence rate, $D_{CSF}$ was scaled so that $D_{CSF} = 100 D_{CSF}\sp{\star} $. The initial values of the diffusion coefficients in the optimization algorithm was $(D_{CSF}\sp{\star}, D_{GM}, D_{WM})=$  $(1, 1, 1)$.

The noise susceptibility was tested by adding a uniform distribution noise term to the observations. The noise term was constructed using the random library in numpy and adjusted so that the noise range becomes $\lbrace -n_{amp} , n_{amp} \rbrace $, with $n_{amp}$ denoted as noise amplitude.

\subsection{Acknowledgements}
This work was performed on the Abel Cluster, owned by the University of Oslo and Uninett\textbackslash Sigma2, and operated by the Department for Research Computing at USIT,the University of Oslo IT-department. \url{http://www.hpc.uio.no/}

@@

The convergence of the minimization is influenced by the values of the regularization parameters $\alpha$ and $\beta$. Therefore the convergence was examined by a systematic study of the reconstruction of the manufactured solution with respect to a wide range of regularization parameters.

@@

Ideally, the approach should have a wide range of parameters in which the reconstruction algorithm yields very similar end-results although the convergence may vary substantially. The evaluation of different regularization parameters was done by solving the optimization with different values of $\alpha$ and $\beta$ and compare the controls with the manufactured solution. The observation times were selected with $t_i \in  \lbrace$ 2.4, 4.8, 7.2, 9.6, 12.0, 14.4, 16.8, 19.2, 21.6, 24.0 $\rbrace$, and the number of time-steps $k$ in the forward problem was set to $10$.

>>>>

\newcommand{\kam}[1]{\textcolor{blue}{#1}}
\newcommand{\fixme}[1]{\textcolor{orange}{#1}}

@@

\begin{abstract}
bla bla 
\end{abstract}
\section{Introduction}
Recent breakthroughs in medicine have linked neurodegenerative diseases like Alzheimer's to insufficient sleep through a novel fluid mechanism called the glymphatic system 
that is responsible for waste clearance. 
In 2011,~\citet{iliff2012paravascular} discovered a novel system within our brain's metabolism system, which changed our understanding of the brain. 
The system involves a new pathway, called the paravascular pathway as it runs in parallel with the vasculature system in a surrounding compartment.
The important novelty in the 2011 findings was that waste from the brain is cleared through the paravascular path, and that a malfunction of this system 
is linked to the development of dementia such as Alzheimer's and Parkinson's diseases. In 2013,~\citet{xie2013sleep} found
that this system is particularly active during sleep.
These mentioned breakthroughs were all performed in mice. In 2018, the action of the glymphatic system was for the first time demonstrated in humans. Further, it was shown that this action differs
between healthy controls and patients with dementia~\citet{ringstad2018brain}.
However, the biomechanical mechanisms behind the glymphatic system is not well understood from a fluid mechanics perspective,
and so far the modeling attempts have mostly failed~\citet{asgari2016glymphatic, holter2017interstitial, smith2017glymphatic}. 
In particular the system seems to facilitate waste transport faster than  
the diffusion \fixme{which diffusion?}, which since the pioneering works of \citet{sykova2008diffusion} 
has been the prevailing paradigm.

@@

In \cite{ringstad2018brain} brain-wide distribution of MRI-contrast was demonstrated during 24 hours after lumbar contrast injection. Brain-wide distribution by diffusion alone was deemed unlikely by the authors (some of which are co-authors of this paper). The argument was based on analytical considerations where it was calculated that 50\% contrast enrichment would occur after 
55 hours using the error function which is valid for planar diffusion. However, 
the surface of the brain is folded and is around five times larger than 
a corresponding surface of a ball with the same volume. Hence, 
a more rigorous modeling attempt is warranted. A complicating factor
is however that the contrast in the surrounding cerebral spinal fluid (CSF) is heterogenous
and changes significantly during the 24 hours of the investigations. 
Furthermore, images were obtained only at a few time-points during the investigations in \cite{ringstad2018brain}. As such, the sparseness in time prevents a direct computation of the diffusion coefficients.

The pioneering work of Sykov{\'a} and Nicholson~\cite{sykova2008diffusion} demonstrated
that diffusion was a governing transport mechanism in the brain and estimated to be on the order
of $1.0e-4 \mathrm{mm\sp{2}/s}$ for large molecules. This is confirmed with diffusion tensor imaging (DTI) where young and healthy subjects have typical diffusion values of around $0.7e-3 \mathrm{mm\sp{2}/s}$ \cite{Helenius194}
while subjects with dementia typically have values of XXX \kam{Geir vet du dette?}. Of course diffusion coefficients, 
measured by DTI at a time-scale of 7-30 minutes, may not be representative for the process on a
longer time-scale, at least if the paravascular pathway plays an important role. The fact that the vasculature system occupies around 3\% of the brain volume 
and the paravascular network is substantially smaller potentially renders it \fixme{what is it here?} invisible at the short time-scales of a DTI investigation 
while it may still be effective on longer time scales.

Our purpose in this paper is to attempt a more rigorous methodology 
for assessing the apparent diffusion process where
the complex geometry of the brain as well as time-scales of hours or days are taken into account.
\fixme{In this paper, we aim to investigate whether we can assess the effective diffusion coefficient on long time-scales (hours or days), by fitting a diffusion model to magnetic resonance imaging (MRI) data.}
Our approach is to solve an optimization problem constrained by a partial differential equation (PDE) using the adjoint method where we have sparse observations on selected time-points, around 10 acquisitions during 24 hours, but through
the complete domain. Hence, the challenges faced from a mathematical point of view 
are that 1) the images are subject to noise, 2) resolution in space is limited to slightly less than (1 mm)$^3$ and 3)
the sparse observations through the time domain.  As such we need to assess the sensitivity of
the approach with respect to important parameters such as the regularization parameters, noise levels and 
time resolution to determine whether this approach is a viable method
to obtain parameters involved time-scales. We remark that the purpose
of this paper is a systematic study of the mathematical challenges and that assessing  
whether clearance is governed by a process that is faster than diffusion is not the topic of the current paper.

@@

Sections 2.1-2.3 describe the details of the imaging relevant for this study. More details can however be
found in e.g. \cite{ringstad2018brain}. Sections 2.4-2.5 describe the solution of PDE constrained optimization problems and
its implementation in FEniCS. The exposition is, however, brief and we refer to XXX for more details.

\subsection{MRI Data}
\begin{figure}
\includegraphics[width=0.95\textwidth]{PatID-68-new-100.png} 
\caption{Shows the percentage intensity increase from baseline at different observation times. The colorbar was restricted to the range $(0,100)$.}
\label{fig1} 
\end{figure}

\begin{figure}
\includegraphics[width=0.95\textwidth]{Zoom-PatID-68.png} 
\caption{Shows the percentage intensity increase from baseline at different observation times in the slice (marked red in the left panel) used in the subsequent analysis. The color bar was restricted to the range $(0,100)$. }
\label{fig2} 
\end{figure}
Figure~\ref{fig1} shows the distribution of tracers during 24 hours, see also~\cite{ringstad2018brain} for further information on the imaging procedure.   
Our data (not all shown) consist of a total of 10 MRI observations, including a baseline MRI taken before tracer was injected. The observation points are distributed over 5 observations within 1-2 hours after injection, a single observation in the timeframes 2-4 hours, 6-9 hours, 24 hours and 48 hours. 
Figure~\ref{fig2} shows the region selected for our computations. 
The software Freesurfer \cite{Dale1999179, FischlLiuDale, spf2007, reuter:robreg10} was used to segment and align each of the observations, which made it possible to estimate voxelwise intensity increase.

@@

concentration. We remark that we use a notation common in medical literature and here two letter symbols are common. Hence, below we will use two letter symbols such as $TE$ and $TR$ to keep the notation consistent with the presentation in \cite{GOWLAND, MPRAGE}.   

@@

The superscripts indicate relaxation time with contrast $T_{1}\sp{c}$ and without contrast $T_{1}\sp{0}$, and $r_1$ is the relaxivity constant for the MRI tracer in a medium. 

@@

Here $T_b$ is known as the echo spacing, $T_a$ is the inversion time, $T_w$ the time delay, $TR$ as the repetition time, $m$ is the number of echo spacings and $T1$ \fixme{Should this be subscripted?} is the longitudinal(spin-lattice) relaxation time for a given medium. The $M_0$ is a calibration constant of the magnetization. The center echo denoted as $n=m/2$ will be the signal that we will consider when estimating MRI tracer. Given Eq.\ref{EQ::SI} we have that the relative intensity increase can be written as 

@@

\frac{S\sp{c}}{S\sp{0}} = \frac{ M_{n}\sp{c} \sin (\theta)}{ M_{n}\sp{0} \sin (\theta) }.

@@

which can be seen in Figure \ref{figuredti}. 

@@

and then estimate the concentration using \eqref{EQ::contrast}. The $T_{1}\sp{0}$ values were obtained by T1 mapping the brain using a MRI sequence known as MOLLI5(3)3 \cite{TAYLOR201667}. This takes into account patients specific characteristic, such as tissue damage. Tissue damage can be observed in the MRI due to a lower intensity in the white matter compared to healthy white matter tissue, thus damaged tissue have different $T_1$ relaxation time. \fixme{How is $r_1$ computed?}

\begin{figure}
\centering
\includegraphics[width=0.70\textwidth]{T1function.png} 
\caption{Shows the  function defined in \eqref{scaledmagnetization} where the white region indicates $T_1$ values for white matter, the grey region indicates $T_1$ values for grey matter, the blue region indicates $T_1$ values for CSF.  }
\label{figuredti} 
\end{figure}

\subsection{Diffusion tensor imaging}
In addition to the T1 and T2 weighted sequences described above we have also obtained diffusion tensor imaging to assess the apparent diffusion 
coefficients on short time-scales. Images are shown in Figure~\ref{figuredti} where the largest diffusion coefficient (shown in 
red in the middle figure) is shown to be around $1.0e-3 mm^2/s$. We remark that we have not included possible anisotropy, shown in 
the right-most image in Figure~\ref{figuredti} and that these images show the apparent diffusion coefficient for free water molecules (18 Da).      
The diffusivity of the Gadovist (600 Da \cite{MGadobutrol}) was estimated to be similar to the diffusion coefficient of Gd-DPTA (550 Da \cite{MGgDPTA}). This is due to the fact that both molecules have similar mass, and based on Stoke-Einstein equation should have similar diffusion coefficients. The free diffusion coefficient for Gd-DPTA was estimated in \citet{GdDPTA-DIFFUSION} to be $3.8e-4\mathrm{mm\sp{2}/s}$.
The fractional anisotropy is defined as 
\begin{equation}
FA\sp{2} =  \frac{3}{2} \frac{ (\lambda_1 - MD )\sp{2} +(\lambda_2 - MD )\sp{2} +(\lambda_3 - MD )\sp{2}}{\lambda\sp{2}_1 + \lambda\sp{2}_2  +\lambda\sp{2}_3 },
\end{equation}
with the mean diffusivity $MD$ defined as 
\begin{equation}
MD = \frac{\lambda_1 +\lambda_2 +\lambda_3 }{3}.
\end{equation}
In these equations $\lambda_i$ denotes the eigenvalues of the diffusion tensor.
\begin{figure}
\centering
\includegraphics[width=0.80\textwidth]{DTI-zoom.png} 
\caption{The left panel show the anatomical map. The middle panel show the apparent diffusion coefficients (ADC) obtained from DTI. The right panel shows the computed fractional anisotropy (FA) from the DTI.}
\label{figuredti} 
\end{figure}

\subsection{Mathematical Model}
In \cite{sykova2008diffusion}, it was shown that the macroscopic diffusion in the brain can be considered a hindered diffusion with an apparent diffusion coefficient. In order to estimate the apparent diffusion coefficient involved in the tracer transportation shown in Figure~\ref{fig1} we assume that the process can be modeled by a diffusion equation. Then we solve a PDE constrained optimization problem where
the tracer distribution, boundary conditions, and apparent
diffusion coefficient are solved for by optimizing with
respect to the observed tracer distribution. As such, enhanced transportation because of effects such as
dissipation would result in an apparent diffusion coefficient larger than that predicted by DTI.  
The objective function was defined as 
\begin{equation}
\min_{D,g} \quad \sum\limits_{i=1}\sp{n} \int\limits_{\Omega} |u(t_i) - u_{obs}(t_i)|\sp{2} \mathrm{d}\Omega + \int\limits_{0}\sp{T} \int\limits_{\partial \Omega_1} \frac{\alpha}{2} | g |\sp{2} + \frac{\beta}{2} \left| \frac{\partial g}{\partial t} \right|\sp{2} \mathrm{d}\Omega \mathrm{d}t 
\label{EQ::objf}
\end{equation}
subject to   
\begin{equation}
\begin{aligned}
\frac{\partial u}{\partial t} &= \nabla \cdot  D \nabla u && \text{in} \qquad \Omega \times \left\lbrace 0 , T \right)  \\
u&=g && \text{on} \qquad \partial\Omega_1  \times \left\lbrace 0 , T \right) 
\end{aligned}
\label{Eq::PDE}
\end{equation}
Here, $u$ is the contrast distribution, $D$ is the apparent diffusion 
coefficient, $g$ is the boundary condition, $\Omega$ is the domain, and $T$ is the final simulation time. \fixme{Should we use $c$ instead of $u$?} We assume that the domain $\Omega$ consists of three sub domains, each with a different diffusion coefficient. We denote the CSF domain as $\Omega_1$, the grey matter as $\Omega_2$ and the white matter as $\Omega_3$. 
The apparent
diffusion constant is assumed to be constant within the CSF, grey and 
white matter but each region may have different values.  
The Dirichlet boundary condition is only applied on the outer facing boundary of the CSF domain, $\partial \Omega_1$. Homogeneous Neumann conditions are applied on the remaining boundaries.
The $\alpha$ and $\beta$ parameters are non-negative regularization parameters 
and $u_{obs}$ are the observations at time-points $t_i$.

The mesh construction was patient-specific and used the MRI of a patient diagnosed with idiopathic normal pressure hydrocephalus (iNPH). The software Freesurfer was used to segment and create the polyhedral surfaces of the white and grey matter. Then the T2 weighted MRI \cite{eidevalnes} was used to segment the CSF compartment surrounding the cerebral and the lateral ventricles. CGAL \cite{cgal:rty-m3-18b} was used to combine the polyhedral surfaces and to construct the mesh. The computational requirement for the resulting mesh was significant, therefore two submeshes was also constructed, see Fig.\ref{Fig::Mesh}.
\begin{figure}
\centering
\includegraphics[scale=0.2]{mesh.png} 
\caption{A) Shows the mesh created from the baseline MR image with 3 domains. B) Shows the mesh created from the baseline MR image with 2 domains. \fixme{Add explaination for which color corresponds to which domain}  }
\label{Fig::Mesh}
\end{figure}
The 3 domain mesh in Fig.\ref{Fig::Mesh} A, consists of 244318 tetrahedral cells and 22057 vertices, while the 2 domain mesh (without the CSF compartment) Fig.\ref{Fig::Mesh} B consists of 335589 tetrahedral cells and 73002 vertices. The dimension of the mesh coordinates are \fixme{missing value} $\mathrm{mm}$.

\subsection{Manufactured Solution}
In order to asses the robustness and accuracy of our strategy and test the dependency of the computed diffusion coefficients $D$ on the
regularization parameters $\alpha$ and $\beta$, we perform a test case with a
known solution.  
The manufactured observations were obtained by forward computation of Eq.\ref{Eq::PDE} with the Dirichlet boundary condition defined as
\begin{equation}
g(t) = 0.3 +0.167t - 0.007t\sp{2} \qquad \text{ for } 0 \leq t \leq 24.
\label{EQ::DIRI}
\end{equation}
The timestep was $dt = 0.02$, and the diffusion coefficients were selected to be \fixme{describe discretisation in time and space}
\begin{equation}
D_{\Omega_1} = 1000.0, \quad D_{\Omega_2} = 4.0, \quad D_{\Omega_3} = 8.0 
\end{equation}  
The magnitude order for $D_{\Omega_1} $ and $D_{\Omega_2}$ were chosen to resemble diffusion coefficient for water in grey and white matter, but the relation between the coefficients were not preserved. The diffusion coefficient in the CSF was chosen given [cite Bryne and Kent], and the initial condition was set to 0 everywhere. \fixme{Check consistency of notation for $D_{\Omega_3}$, for example in next section it was called $D_{CSF}$} %and the initial condition was set to 1.0 in $\Omega_1$ and 0 elsewhere.
The manufactured forward computation gave a total of 120 possible observation times. \fixme{Should this be 1200? 24/0.02 = 1200}.

\section{Implementation}
The tracer concentration was estimated in a preprocessing step, using the parameters obtained from the T1 \fixme{should this be subscripted? Otherwise unclear.} map, MPRAGE MRI protocol \cite{eidevalnes} and the value for $r_1$ found in \cite{pmid16230904} \fixme{rewritten sentence, please check}. In the computation, the function Eq.\ref{scaledmagnetization} was computed for $ T_1\in \lbrace 200, 4000 \rbrace$ creating a lookup table. The lookup table was utilized with the baseline intensity increase to estimate $T_1\sp{c}$, and then the concentration was computed using Eq.\ref{EQ::contrast}.

@@

The solver for \eqref{Eq::PDE} was implemented using the FEniCS project (v.2017.2), using backwards Euler time discretization and first order continuous Galerkin finite elements in space. The resulting linear problems were solved using GMRES.
The module dolfin-adjoint \cite{farrell2013automated, funke2013framework} was used to automatically derive and solve the adjoint problem, and to solve the PDE constrained optimization problem with the L-BFGS-B algorithm \cite{LBFGSB1, LBFGSB2}. The optimization was stopped when the $L^\infty$-norm of the projected gradient of the objective functional dropped below $1.0e-1$.

The observations $u_{obs}$ have fixed time-points. However, these time-points may not coincide  with the time discretization in the forward problem $t_j$. Therefore, we linearly interpolated the computed solutions onto the observation time points using the formula
\begin{equation}
\label{observation:interpolation}
u(t) \approx \frac{\Delta t}{dt} u(t_{j-1}) + \frac{dt - \Delta t }{dt} u(t_{j}), \quad t \in \lbrace t_{j-1}, t_j \rbrace
\end{equation}
with $\Delta t = t_j-t$ and $dt = t_j - t_{j-1}$.

The implementation used the first observation as initial conditions of Eq.\ref{Eq::PDE}. Then for each time-step, the next observation was used as the initial value for boundary control $g$. In order to increase the convergence speed of the optimization, $D_{CSF}$ was scaled so that $D_{CSF} = 100 D_{CSF}\sp{\star} $. The initial values of the diffusion coefficients in the optimization algorithm were $(D_{CSF}\sp{\star}, D_{GM}, D_{WM})=$  $(1, 1, 1)$. \fixme{Different $D$ notation compare to previous sections}

The noise susceptibility was tested by adding a uniform \fixme{pointwise?} distribution noise to the observations. The noise term was constructed using the random library in numpy and adjusted so that the noise range was $\lbrace -n_{amp} , n_{amp} \rbrace $, with $n_{amp}$ denoted as noise amplitude.

\subsection{Acknowledgements}
The computational experiments were performed on the Abel Cluster, owned by the University of Oslo and Uninett\textbackslash Sigma2, and operated by the Department for Research Computing at USIT,the University of Oslo IT-department. \url{http://www.hpc.uio.no/}

@@

The convergence as well as the result of the minimization is influenced by the values of the regularization parameters $\alpha$ and $\beta$. Therefore the convergence was examined by a systematic study of the reconstruction of the manufactured solution with respect to a wide range of regularization parameters.

@@

Ideally, our approach should have a wide range of parameters in which the reconstruction algorithm yields very similar end-results although the convergence may vary substantially. The evaluation of different regularization parameters was done by solving the optimization with different values of $\alpha$ and $\beta$ and comparing the computed $D$ and $g$ values with the manufactured solution. The observation times were set to $\lbrace$ 2.4, 4.8, 7.2, 9.6, 12.0, 14.4, 16.8, 19.2, 21.6, 24.0 $\rbrace$, and the number of time-steps $k$ in the forward problem was set to $10$.
